---
title: "Alternative to Maximum Likelihood for two-parameter Matern Fields"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Fitting a Matérn covariance to a (possibly  incomplete) lattice observation 

The   `CGEMEV` R package provides tools for simulating realizations of a stationary isotropic Gaussian process when the correlation belongs to the common Matérn family with known smoothness index $\nu \geq 0$ and for estimating  the correlation  range (also called ''decorrelation length'') from one realization on a (possibly incomplete) lattice. 
The variance of the field is simply estimated by the empirical variance that will be denoted $b_{EV}$.
Presently the missing regions can be chosen to be several disks.

Three main functions are used: `gaussian.matern()`, `simulate()`
and `fsai11Precond.GEevalOnThetaGrid()`,  and a fourth function `grid.domain()` is required to precompute preconditioning sparse matrices.

These fonctions can be applied to a quite large grid even on a laptop (for example 512x512, provided the ''extension factor'' required for simulation, see below, is not too big). Indeed quite fast 
computation of the quadratic form which occurs in the estimating equation is possible by using a conjugate-gradient (CG) solver preconditioned by a classical factored sparse approximate inverse (FSAI) preconditioning, since the matrix-vector product, required in each CG iteration,  can be obtained via FFT from the standard embedding of the correlation matrix  in a circulant  matrix.

## Contents

- [Setting the probabilistic model](#Setting-the-probabilistic-model)
- [Simulating one realization](#Simulating-one-realisation)
- [Plotting (and saving) several realizations](#plotting-(and-saving)-several-realizations)
- [Setting the uncomplete lattice](#setting-th-uncomplete-lattice)
- [Plotting data](#Plotting-data)
- [Plotting the estimating function](#plotting-the-estimating-function)
- [Estimating theta and the micro-ergodic parameter](#estimating-theta-and-the-micro-ergodic-parameter)

## Setting the probabilistic model

In this first version of the  `CGEMEV` package, for simplicity, we restrict the spatial domain to be the unit square (0,1)X(0,1). For the example here, the simulations and the choice of observed sites are done using a grid 48x48 partitioning this domain.
```{r }
library(CGEMEV)
n1grid<-48
# gaussian matern creation
gm <- gaussian.matern(grid.size=n1grid,smoothness=0.5,range=0.5,factor=2)
```
NB: in the previous setting, ''factor=2'' specifies the required extension factor of the observation domain. Indeed for this example the choice ''factor=1'' would entail (when calling `simulate()`) the message "FFT of covariance has negative values" which means that generating a realization via the classical embedding method (which doubles each length of the considered rectangular domain) would not work.

## Simulating one realization

This is simply:
```{r }
set.seed(321)  # so that it is reproducible #
simulate(gm)
```

## Plotting (and saving) several realizations

We can plot (and save), the previous realization and, for example, 8 further realizations:


```{r}
fullLattice.nineZs<- array(NA,c(n1grid*n1grid,9))
set.panel(2,3)
plot(gm)
fullLattice.nineZs[,1]<-gm$look[1:gm$n1,1:gm$n1]
ut <- system.time(
for (indexReplcitate in 2:9){
  set.seed(321+indexReplcitate)
  simulate(gm)
  plot(gm)
  fullLattice.nineZs[,indexReplcitate]<-gm$look[1:gm$n1,1:gm$n1]
  # image(x1,x2,matrix(Ztrue, gm1over2range0p5$n1, gm1over2range0p5$n1),asp=1)}
})
```

The following timing is for a MacBookPro3,1 Intel Core 2 Duo 2.6GHh  :
```{r}
ut   # for the simulation of 8 realisations :
```


## Setting the uncomplete lattice

Let us now define the regions  (actually 2 disks) where the observations will be missing, and precompute the preconditioning matrix:
```{r}
# md=missing.domains
ex1.md <- list(
  list(center=c(0.67841,0.67841),radius=0.17841),
	list(center=c(0.278412, 0.228412),radius=0.071365)
)
# gd=grid.domain
print(system.time(ex1WithN1eq48And2missindDisks.gd <- grid.domain(ex1.md,n1grid)))
```

## Plotting data

TODO


## Plotting the estimating function

Consider the first one of the above realizations, and the naive variance estimator $b_{\rm EV}$:
```{r}
# only observed outside the disks:
#z <- gm$look[1:gm$n1,1:gm$n1][!ex1WithN1eq48And2missindDisks.gd$missing.sites]
indexReplcitate <- 1
z <- fullLattice.nineZs[,indexReplcitate][!ex1WithN1eq48And2missindDisks.gd$missing.sites]
#
(bEV  <- mean(z**2))
```
For this realization, let us give the whole output of the function `fsai11Precond.GEevalOnThetaGrid()` :
```{r}
# choosing the inverse-range values (the theta's) at which the estimating funcion is computed:
candidateThetas1DGrid <- 1/gm$range * 10**seq(-1.1,1.1,,15)

(out <- fsai11Precond.GEevalOnThetaGrid(z,candidateThetas1DGrid,nu=gm$smoothness,                          
grid.domain=ex1WithN1eq48And2missindDisks.gd,tolPGC=1e-03)
)
#
#
set.panel(2,3)
{plot(candidateThetas1DGrid, out$values, type="l", 
                 col=1, lty=2,log="x")
    abline(h= bEV)
}
ut <- system.time(
for (indexReplcitate in 2:9){
  z <- fullLattice.nineZs[,indexReplcitate][!ex1WithN1eq48And2missindDisks.gd$missing.sites]
  bEV  <- mean(z**2)
  out <- fsai11Precond.GEevalOnThetaGrid(z,candidateThetas1DGrid,
        nu=gm$smoothness,grid.domain=ex1WithN1eq48And2missindDisks.gd,tolPGC=1e-03)
  plot(candidateThetas1DGrid, out$values, type="l", 
                 col=1, lty=2,log="x")
    abline(h= bEV)
})
```

Timing  for a MacBookPro3,1 Intel Core 2 Duo 2.6GHh :
```{r}
ut   # for computing the estimating equation for 8 realisations :
```

## Estimating theta and the micro-ergodic parameter

TODO
#
